---
title: "Script to create figures related to the immunopeptidomics data"
output: html_document
author: "Leron Kok"
---

# Pre-processing
## Declare libraries
```{r}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(patchwork)
  library(Peptides)
  library(ggpubr)
  library(grid)
  library(stringi)
  library(RColorBrewer)
  library(scales)
})
```

## Load required data and define directories and files
```{r}
utils::globalVariables(c("protein", "detected", "order_ids", "y_val",
                         "fill_group"))
source("load_tables.R")
source("load_peptideatlas_data.R")
source("load_plot_aesthetics.R")

processed_dir <- here("processed")
plot_dir <- file.path(processed_dir, "plots")

# GTEX FPKM mean expression data (excluding testis)
gtex_fpkm_file <- file.path(raw_dir, "GTEX_FPKMmean_expression.txt")
gtex_data <- read.csv(gtex_fpkm_file)

# List of genes associated with cancer from the Cancer Gene Census
cancer_census_file <-
  file.path(raw_dir, "Census_allThu_Jan_4_14_08_58_2024.csv")

# Canonical proteins detected on HLA-I
prot_can_hlai <- pep_can %>%
  ungroup() %>%
  semi_join(peptide_hla %>%
              filter(hla_class == "I"), "peptide") %>%
  distinct(protein_accession) %>%
  semi_join(can_id, "protein_accession")

# ncORFs detected on HLA-I (one peptide may be linked to multiple ncORFs)
prot_nonc_hlai <- pep_nonc_long %>%
  ungroup() %>%
  semi_join(peptide_hla %>%
              filter(hla_class == "I"), "peptide") %>%
  distinct(protein_accession)

# HLA-I detectability of ncORF and CDS
seq_stats <- can_nonc_seq %>%
  # Filter out protein sequences with Selenocysteine (not recognized by the
  #  Peptides package)
  filter(!str_detect(protein, "U")) %>%
  mutate(
    pro_cat = ifelse(str_starts(protein_accession, "CONTRIB_GENCODE_"),
      "Non-canonical", "Canonical"
    ),
    len = nchar(protein)
  ) %>%
  # Filter for minimum length of 16 amino acids (used for ncORF list creation)
  filter(len >= 16) %>%
  left_join(
    bind_rows(prot_can_hlai, prot_nonc_hlai) %>%
      mutate(detected = "Detected"), "protein_accession"
  ) %>%
  left_join(
    orf_sequences %>% distinct(protein_accession, orf_biotype),
    "protein_accession"
  ) %>%
  replace_na(list(detected = "Undetected", orf_biotype = "CDS"))
```

## Define functions
```{r}
get_stats <- function(df) {
  # Determine the length, hydrophobicity and isoelectric point for proteins
  #
  # Args:
  #   df: dataframe with protein column
  #
  # Returns:
  #   Same as input, with the added columns len, hydro and pi
  df %>% mutate(
    len = nchar(protein),
    hydro = hydrophobicity(protein, "KyteDoolittle"),
    pi = pI(protein)
  )
}

perform_wtest <- function(df) {
  # Perform wilcox test
  #
  # Args:
  #   df: dataframe with the columns y_val and detected
  #
  # Returns:
  #   p-value outputted by the wilcox test
  w_test <- wilcox.test(y_val ~ detected,
    data = df %>% mutate(detected = factor(
      detected,
      levels = c("Detected", "Undetected")
    ))
  )
  p_val <- w_test$p.value
  return(p_val)
}

format_pval <- function(p_val) {
  # Process raw p-value to a formatted string
  #
  # Args:
  #   p_val: raw p-value as a number
  #
  # Returns:
  #   string with the formatted p-value
  p_val <- case_when(
    p_val >= 0.05 ~ "p = n.s.",
    p_val >= 0.010 ~ paste0("p = ", round(p_val, 3)),
    p_val == 0 ~ "p = 0",
    .default = paste0("p = ", format(p_val, scientific = TRUE, digits = 4) %>%
                        str_replace("e", "\u00D710"))
  )
  return(p_val)
}

get_pvals_holms <- function(df) {
  # Determine p-values with Bonferroni–Holm correction
  #
  # Args:
  #   df: dataframe with the column detected and any number of columns with
  #    test values
  #
  # Returns:
  #   dataframe with for each of the test columns a formatted p-value
  df %>%
    pivot_longer(-detected, values_to = "y_val") %>%
    split(~name) %>%
    sapply(perform_wtest) %>%
    p.adjust(method = "holm") %>%
    sapply(format_pval) %>%
    enframe()
}

create_violin <- function(violin_data, y_label, p_val, y_lim,
                          n_colors = c(1:8), log = FALSE, pseudo_log = FALSE,
                          x_label = element_blank(), y_breaks = waiver()) {
  # Create violin plots split by detected and undetected ORF characteristics
  #
  # Args:
  #   violin_data: dataframe with the data to plot
  #   y_label: string with the y-axis label
  #   p_val: p-value to be shown in the plot
  #   y_lim: limits of the y-axis
  #   n_colors: array with indices of colors to use (from v_colors below)
  #   log: boolean indicating whether to use a logarithmic y-scale,
  #    default False
  #   pseudo_log: boolean indicating whether to use a pseudo-log scale,
  #    default False
  #   x_label: string with the x-axis label, default no label
  #   y_breaks: array indicating y-break label positions, default is ggplot
  #    defaults
  #
  # Returns:
  #   ggplot object with the violin plot
  counts <- violin_data %>%
    dplyr::count(detected) %>%
    mutate(n = format(n, big.mark = ",")) %>%
    deframe()
  # Process p-value to be shown in plot
  p_label <- grobTree(textGrob(p_val,
    x = 0.5, y = 1, just = c("center", "top"),
    gp = gpar(col = "black", fontsize = 7)
  ))
  # Standard colors to be used for plots
  v_colors <- c(
    blue_col[c(1, 4)], green_col[c(1, 4)], purple_col[c(1, 4)],
    pink_col[c(1, 4)]
  )[n_colors]
  v_plot <- ggplot(violin_data, aes(order_ids, y = y_val, fill = fill_group)) +
    geom_violin(trim = TRUE, show.legend = FALSE) +
    geom_boxplot(fill = "transparent", outlier.shape = NA) +
    scale_fill_manual(values = v_colors) +
    scale_x_discrete(
      name = ifelse(x_label == "", element_blank(), x_label),
      labels = c(
        paste0(
          "Detected\n(n = ",
          counts[["Detected"]], ")"
        ),
        paste0(
          "Undetected\n(n = ",
          counts[["Undetected"]], ")"
        )
      )
    ) +
    annotation_custom(p_label)
  if (log) {
    v_plot <- v_plot + scale_y_log10(
      name = y_label, limits = y_lim,
      breaks = y_breaks
    )
  } else if (pseudo_log) {
    v_plot <- v_plot + scale_y_continuous(
      trans = scales::pseudo_log_trans(base = 10),
      name = y_label, limits = y_lim, breaks = y_breaks
    )
  } else {
    v_plot <- v_plot + scale_y_continuous(
      name = y_label, limits = y_lim,
      breaks = y_breaks
    )
  }
  return(v_plot)
}

create_dis_plot <- function(plot_data, x_name, plot_colors, x_breaks = waiver(),
                            y_breaks = waiver(), y_name = "Density") {
  # Create density plot split by cancer sample detectability of peptides
  #
  # Args:
  #   plot_data: dataframe with the data to plot
  #   x_name: string with the x-axis label
  #   plot_colors: colors to be used in the plot
  #   x_breaks: array indicating y-break label positions, default is ggplot
  #    defaults
  #   y_breaks: array indicating y-break label positions, default is ggplot
  #    defaults
  #   y_name: string with the y-axis label, default is "Density"
  #
  # Returns:
  #   ggplot object with the density plot
  plot_data +
    geom_density(
      alpha = .4, adjust = 1.5, linewidth = 0.5,
      show.legend = FALSE
    ) +
    scale_x_continuous(expand = c(0, 0), breaks = x_breaks, name = x_name) +
    scale_y_continuous(expand = c(0, 0), breaks = y_breaks, name = y_name) +
    scale_color_manual(values = plot_colors)
}
```

# Figure 2 - HLA figures
## Figure 2E-H - Compare orf categories
```{r}
# Determine certain characteristics for each ORF
orf_pep_stats <- pep_nonc %>%
  rowwise() %>%
  # Determine percentage of ncORF sequence covered by detected peptides
  mutate(cov = paste0(start_loc:end_loc, collapse = ",")) %>%
  group_by(orf_name) %>%
  summarise(n_pep = n(), cov = paste0(cov, collapse = ",")) %>%
  rowwise() %>%
  mutate(n_cov = cov %>%
           str_split(",", simplify = TRUE) %>%
           as.numeric() %>%
           na.omit() %>%
           unique() %>%
           length()) %>%
  select(-cov) %>%
  full_join(orf_sequences, "orf_name") %>%
  replace_na(list(n_pep = 0, n_cov = 0)) %>%
  mutate(
    orf_len = nchar(protein),
    perc_cov = n_cov / orf_len,
    cov_cat = cut(perc_cov,
      breaks = seq(0, 1, 0.1),
      labels = sapply(seq(0, 90, 10), function(i) {
        paste0(i, "-", i + 10, "%")
      })
    )
  ) %>%
  make_biotype_levels() %>%
  ungroup()

# Determine number of ncORFs and peptides per biotype
orf_pep_stats_count <- orf_pep_stats %>%
  group_by(orf_biotype) %>%
  summarise(n_orf = sum(n_pep >= 1), n_pep = sum(n_pep)) %>%
  pivot_longer(-orf_biotype) %>%
  mutate(name = factor(name, levels = c("n_pep", "n_orf")))

# Get Riboseq quality from supplementary table
orf_quality <- hla_orf_table %>%
  filter(n_peptides_in_entry >= 2) %>%
  mutate(
    is_sufficient = Manual.inspection.of.GWIPS.Viz.RiboSeq.Data !=
      "Insufficient",
    study = ifelse(str_detect(identifier, "norep"), "1", ">1"),
    study = factor(study, levels = c("1", ">1"))
  ) %>%
  group_by(study) %>%
  dplyr::count(is_sufficient)

figure_2e <- ggplot(
  orf_pep_stats_count,
  aes(y = value, x = orf_biotype, fill = name)
) +
  geom_bar(position = position_dodge(), stat = "identity",
           show.legend = FALSE) +
  scale_x_discrete(name = element_blank(), expand = c(0, 0)) +
  scale_y_continuous(name = "Number of detections", expand = c(0, 0)) +
  scale_fill_manual(values = green_col[c(6, 1)]) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

figure_2f <- ggplot(orf_pep_stats %>%
                      dplyr::count(n_pep) %>%
                      filter(n_pep > 0), aes(y = n, x = n_pep)) +
  geom_bar(
    position = "stack", stat = "identity", show.legend = FALSE,
    fill = green_col[1]
  ) +
  scale_x_continuous(
    name = "Number of distinct peptides", expand = c(0, 0),
    breaks = 1:16
  ) +
  scale_y_continuous(name = "Number of non-canonical ORFs", expand = c(0, 0)) +
  geom_text(aes(label = n), size = 6 / .pt)

figure_2g <-
  ggplot(
    orf_pep_stats %>%
      filter(perc_cov > 0) %>%
      mutate(n_pep = ifelse(n_pep > 1, "2", "1")),
    aes(perc_cov, orf_len, color = n_pep)
  ) +
  geom_point(show.legend = FALSE) +
  geom_smooth(
    method = "loess", formula = y ~ x, se = TRUE,
    mapping = aes(color = n_pep), linewidth = 1,
    show.legend = FALSE
  ) +
  scale_x_continuous(
    name = "Percentage aa covered", expand = c(0, 0),
    limits = c(0, 1), breaks = c(0, .25, .5, .75, 1),
    labels = c("0%", "25%", "50%", "75%", "100%")
  ) +
  scale_y_continuous(
    name = "ncORF length (aa)", limits = c(0, NA),
    expand = c(0, 0)
  ) +
  scale_color_manual(
    name = element_blank(), values = c(green_col[1], blue_col[4]),
    labels = c("Detected by single peptide", "Detected by multiple peptides")
  )

figure_2h <- ggplot(orf_quality, aes(y = n, x = study, fill = is_sufficient)) +
  geom_bar(position = "stack", stat = "identity", show.legend = FALSE) +
  scale_x_discrete(
    name = "Number of studies\nin which ncORF\nwas detected",
    expand = c(0, 0)
  ) +
  scale_y_continuous(name = "Number of non-canonical ORFs", expand = c(0, 0)) +
  scale_fill_manual(values = c(grey_col[7], green_col[1]))

figure_2eh <- figure_2e + figure_2f + figure_2g + figure_2h +
  plot_layout(nrow = 1, widths = c(1.6, 2.6, 1.4, 1))

ggsave("Figure_2EH.svg",
  device = "svg", path = plot_dir, figure_2eh,
  width = 2500, height = 700, units = "px"
)

# Statistical test for figure 2G - mean length of ncORFs covered by 1 and > 1
#  peptide
wilcox.test(
  data = orf_pep_stats %>%
    filter(perc_cov > 0) %>%
    mutate(n_pep = ifelse(n_pep > 1, "2", "1")),
  orf_len ~ n_pep
)$p.value

orf_pep_stats %>%
  filter(perc_cov > 0) %>%
  mutate(n_pep = ifelse(n_pep > 1, "2", "1")) %>%
  group_by(n_pep) %>%
  summarize(mean_len = mean(orf_len), .groups = "drop")
```

# Figure 3 - Detection determinants
## Figure 3A - Undetected vs detected
```{r}
# Determine pi, hydrophobicity and length for full CDS/ncORF sequences, and
#  for the first and last 30% of the sequences
seq_stats_properties <- seq_stats %>%
  dplyr::rename(full_len = len) %>%
  mutate(to_duplicate = ifelse(pro_cat != "Canonical", 3, 1)) %>%
  uncount(to_duplicate) %>%
  group_by(protein_accession) %>%
  mutate(
    pro_cat = case_when(
      row_number() == 1 ~ paste0(pro_cat, "__full"),
      row_number() == 2 ~ paste0(pro_cat, "__start"),
      row_number() == 3 ~ paste0(pro_cat, "__end")
    ),
    protein = case_when(
      row_number() == 1 ~ protein,
      # Add 0.001 to round to ensure .5 is always rounded up
      row_number() == 2 ~ str_sub(
        protein, 2,
        round(full_len * 0.3 + 0.001) + 1
      ),
      row_number() == 3 ~ str_sub(
        protein,
        full_len - round(full_len * 0.3 + 0.001) + 1,
        full_len
      )
    )
  ) %>%
  get_stats() %>%
  group_by(protein_accession) %>%
  filter(len >= 16) %>%
  select(-protein, -full_len) %>%
  ungroup() %>%
  mutate(
    # Order ids is used for ordering the violin plots
    order_ids = ifelse(detected == "Detected", "1", "2"),
    fill_group = paste0(pro_cat, "__", detected)
  )

l_lab <- "Sequence\nlength (aa)"
h_lab <- "Hydrophobicity\n(Kyte-Doolittle)"
p_lab <- "Isoelectric\npoint"
v_cats <- c("Non-canonical__full", "Canonical__full")

# Determine Bonferroni–Holm corrected p-values (corrected separately for
#  CDS and ncORFs)
p_vals_violin <-
  bind_rows(
    seq_stats_properties %>%
      filter(pro_cat == v_cats[1]) %>%
      select(pi, hydro, len, detected) %>%
      get_pvals_holms() %>%
      mutate(cat = "noncanonical"),
    seq_stats_properties %>%
      filter(pro_cat == v_cats[2]) %>%
      select(pi, hydro, len, detected) %>%
      get_pvals_holms() %>%
      mutate(cat = "canonical")
  ) %>%
  pivot_wider(names_from = cat, values_from = value) %>%
  column_to_rownames("name")

figure_3a1 <- lapply(1:2, function(i) {
  cat <- v_cats[i]
  limits <- list(c(NA, NA), c(NA, NA))
  plot_data <- seq_stats_properties %>%
    filter(pro_cat == cat) %>%
    mutate(y_val = len)
  p_val <- p_vals_violin["len", i]
  create_violin(plot_data, l_lab, p_val, limits[[i]],
    n_colors = c((4 * i - 1):(4 * i)), log = TRUE
  )
})
figure_3a2 <- lapply(1:2, function(i) {
  cat <- v_cats[i]
  plot_data <- seq_stats_properties %>%
    filter(pro_cat == cat) %>%
    mutate(y_val = hydro)
  p_val <- p_vals_violin["hydro", i]
  create_violin(plot_data, h_lab, p_val, c(-4.1, 2.5),
    n_colors = c((4 * i - 1):(4 * i))
  )
})
figure_3a3 <- lapply(1:2, function(i) {
  cat <- v_cats[i]
  plot_data <- seq_stats_properties %>%
    filter(pro_cat == cat) %>%
    mutate(y_val = pi)
  p_val <- p_vals_violin["pi", i]
  create_violin(plot_data, p_lab, p_val, c(2.9, 14),
    n_colors = c((4 * i - 1):(4 * i))
  )
})

figure_3a <- lapply(1:6, function(i) {
  one_plot <- c(figure_3a1, figure_3a2, figure_3a3)[[i]]
  if (i <= 4) {
    one_plot <- one_plot + remove_x
  }
  if (i %in% c(4, 6)) {
    one_plot <- one_plot + remove_y
  }
  if (i == 2) {
    one_plot <- one_plot + theme(axis.title.y = element_blank())
  }
  return(one_plot)
}) %>% wrap_plots(figure_3a_plots, ncol = 2)

ggsave("Figure_3A.svg",
  device = "svg", path = plot_dir, figure_3a,
  width = 1250, height = 1750, units = "px"
)
```

## Figure 3B - Hydrophobicity end of sequence
```{r}
# Using a sliding frame of 15 amino acids, determine the average hydrophobicity
#  for the last 100 amino acids for canonical proteins and ncORFs
seq_stats_hydro <- seq_stats %>%
  mutate(
    protein = stri_reverse(protein),
    duplicate = 101
  ) %>%
  uncount(duplicate) %>%
  group_by(protein_accession) %>%
  mutate(
    start = row_number(), end = start + 14,
    protein = str_sub(protein, start, end),
    pro_len = nchar(protein),
    c_offset = -(start - 1),
    hydro = hydrophobicity(protein, "KyteDoolittle")
  ) %>%
  # Only keep average hydrophobicity if it is based on 15 amino acids
  filter(pro_len >= 15)

# Group average hydrophobicity by biotype and add confidence intervals
seq_stats_hydro_grouped <- seq_stats_hydro %>%
  group_by(orf_biotype, c_offset) %>%
  mutate(
    low_conf = t.test(hydro)$conf.int[1],
    high_conf = t.test(hydro)$conf.int[2]
  ) %>%
  summarise(
    hydro = mean(hydro), low_conf = first(low_conf),
    high_conf = first(high_conf)
  ) %>%
  filter(!orf_biotype %in% c("doORFs", "Processed transcript ORFs"))

# Factor orf biotype by line order on right side of graph
seq_stats_hydro_grouped_order <- seq_stats_hydro_grouped %>%
  filter(c_offset == 0) %>%
  arrange(-hydro) %>%
  pull(orf_biotype)

seq_stats_hydro_grouped <- seq_stats_hydro_grouped %>%
  mutate(orf_biotype = factor(orf_biotype,
    levels = seq_stats_hydro_grouped_order
  ))

figure_3b <- ggplot(
  seq_stats_hydro_grouped,
  aes(x = c_offset, y = hydro, group = orf_biotype)
) +
  geom_linerange(aes(ymin = low_conf, ymax = high_conf, color = orf_biotype),
    alpha = 0.5, linewidth = 0.3
  ) +
  geom_point(mapping = aes(color = orf_biotype), size = .2) +
  geom_smooth(
    method = "loess", formula = y ~ x, se = FALSE,
    mapping = aes(color = orf_biotype), linewidth = .7
  ) +
  scale_x_continuous(
    expand = c(0, 0),
    name = "Position relative to C-terminus"
  ) +
  scale_y_continuous(
    limits = c(-1, 0.5), expand = c(0, 0),
    name = "Hydrophobicity (Kyte-Doolittle)"
  ) +
  scale_color_manual(
    values = brewer.pal(6, "Dark2"),
    name = "ORF/protein category"
  )

ggsave("Figure_3B.svg",
  device = "svg", path = plot_dir, figure_3b,
  width = 1250, height = 1250, units = "px"
)
```

## Figure 3C - Peptide location
```{r}
# Determine peptide start and end offsets for canonical and noncanonical
#  peptides
pep_pos_can <- pep_can %>%
  inner_join(can_id, "protein_accession") %>%
  inner_join(can_nonc_seq, "protein_accession") %>%
  semi_join(peptide_hla %>%
              filter(hla_class == "I"), "peptide") %>%
  distinct(protein, peptide, start_loc, end_loc) %>%
  mutate(
    len = nchar(protein),
    start_offset = start_loc - 1,
    end_offset = -(len - end_loc)
  )

pep_pos_nonc <- pep_nonc %>%
  semi_join(peptide_hla %>%
              filter(hla_class == "I"), "peptide") %>%
  distinct(protein, peptide, start_loc, end_loc) %>%
  mutate(
    len = nchar(protein),
    start_offset = start_loc - 1,
    end_offset = -(len - end_loc)
  )

# Determine max y-axis range for plots
can_lim <- c(0, pep_pos_can %>% filter(end_offset == 0) %>% nrow())
nonc_lim <- c(0, pep_pos_nonc %>% filter(end_offset == 0) %>% nrow())

# Determine fold change and fisher exact test p-values for start and end offset
fold_changes <- bind_rows(
  pep_pos_can %>% mutate(cat = "can"),
  pep_pos_nonc %>% mutate(cat = "nonc")
) %>%
  ungroup() %>%
  select(cat, start_offset, end_offset) %>%
  pivot_longer(-cat) %>%
  filter((name == "start_offset" & value <= 1) |
           (name == "end_offset" & value >= -1)) %>%
  dplyr::count(cat, name, value) %>%
  mutate(value = factor(value, c(0, 1, -1))) %>%
  arrange(value) %>%
  group_by(cat, name) %>%
  mutate(fc = first(n) / last(n)) %>%
  ungroup()

fisher_tests <- sapply(c("start_offset", "end_offset"), function(i) {
  fold_changes %>%
    filter(name == i) %>%
    select(n, cat, value) %>%
    pivot_wider(values_from = n, names_from = cat) %>%
    select(-value) %>%
    fisher.test() %>%
    .$p.value
})

can_start <- ggplot(
  pep_pos_can %>% filter(start_offset <= 25),
  aes(x = start_offset)
) +
  geom_histogram(
    binwidth = 1, color = "white", linewidth = 0.2,
    fill = pink_col[1]
  ) +
  scale_x_continuous(
    name = element_blank(),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "Number of\ncanonical peptides",
    expand = c(0, 0),
    n.breaks = 3,
    labels = label_number(scale_cut = cut_short_scale()),
    limits = can_lim
  ) +
  theme(axis.ticks.x = element_blank(), axis.text.x = element_blank())

can_end <- ggplot(
  pep_pos_can %>% filter(end_offset >= -25),
  aes(x = end_offset)
) +
  geom_histogram(
    binwidth = 1, color = "white", linewidth = 0.2,
    fill = pink_col[1]
  ) +
  scale_x_continuous(
    name = element_blank(),
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = element_blank(),
    expand = c(0, 0),
    limits = can_lim
  ) +
  theme(axis.ticks = element_blank(), axis.text = element_blank())

nonc_start <- ggplot(
  pep_pos_nonc %>% filter(start_offset <= 25),
  aes(x = start_offset)
) +
  geom_histogram(
    binwidth = 1, color = "white", linewidth = 0.2,
    fill = green_col[1]
  ) +
  scale_x_continuous(
    name = "Position relative to N-terminus (aa)",
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = "Number of\nncORF peptides",
    expand = c(0, 0),
    n.breaks = 3,
    limits = nonc_lim
  )

nonc_end <- ggplot(
  pep_pos_nonc %>% filter(end_offset >= -25),
  aes(x = end_offset)
) +
  geom_histogram(
    binwidth = 1, color = "white", linewidth = 0.2,
    fill = green_col[1]
  ) +
  scale_x_continuous(
    name = "Position relative to N-terminus (aa)",
    expand = c(0, 0)
  ) +
  scale_y_continuous(
    name = element_blank(),
    expand = c(0, 0),
    limits = nonc_lim
  ) +
  theme(axis.ticks.y = element_blank(), axis.text.y = element_blank())

figure_3c <- (can_start + can_end) / (nonc_start + nonc_end)

ggsave("Figure_3C.svg",
  device = "svg", path = plot_dir, figure_3c,
  width = 1100, height = 800, units = "px"
)
```

## Figure 3D - Expression
```{r}
# Determine mean fpkm per gene
gtex_df <- gtex_data %>%
  dplyr::rename(gene_id = X) %>%
  pivot_longer(-gene_id) %>%
  group_by(gene_id) %>%
  summarise(mean_fpkm = mean(value))

# Link mean fpkm to ncORF genes
orf_gtex <- seq_stats %>%
  inner_join(orf_sequences %>%
               select(protein_accession, gene_id), "protein_accession") %>%
  inner_join(gtex_df, "gene_id") %>%
  mutate(y_val = mean_fpkm) %>%
  mutate(
    order_ids = ifelse(detected == "Detected", "1", "2"),
    fill_group = paste0(pro_cat, "__", detected)
  )

# Number of missing ncORFs genes in GTEx
orf_sequences %>%
  anti_join(orf_gtex, "protein_accession") %>%
  nrow()

# Mean FPKM statistics
orf_gtex %>%
  group_by(detected) %>%
  summarise(mean_y = mean(y_val))

p_val_gtex <- perform_wtest(orf_gtex) %>% format_pval()

figure_3d <- create_violin(orf_gtex %>% filter(y_val > 0), "Mean FPKM",
  p_val_gtex, c(0, 1600),
  n_colors = c(3, 4),
  pseudo_log = TRUE,
  y_breaks = c(0, 10, 250, 500, 1000, 2000)
)

ggsave("Figure_3D.svg",
  device = "svg", path = plot_dir, figure_3d,
  width = 625, height = 600, units = "px"
)
```

# Figure S6 - HLA class and ncORF peptide origin
## Figure S6A - Peptide classification
```{r}
# Determine number of CDS and ncORF per length and HLA class category
barplot_data <- peptide_hla %>%
  group_by(peptide) %>%
  mutate(hla_class = ifelse(n_distinct(hla_class) > 1, "Both", hla_class)) %>%
  ungroup() %>%
  distinct(peptide, hla_class, pep_type) %>%
  mutate(
    pep_len = nchar(peptide),
    pep_len_cat = case_when(
      pep_len < 8 ~ "7",
      pep_len > 12 ~ "13-35",
      TRUE ~ "8-12"
    )
  ) %>%
  dplyr::count(hla_class, pep_type, pep_len_cat)

# Process barplot_data to include percentages and colors per bar
## Category 1: Sample count
cat1 <- peptide_hla %>%
  distinct(dataset_id, ms_name, hla_class) %>%
  dplyr::count(hla_class) %>%
  mutate(cat = 1, pep_type = "all") %>%
  dplyr::rename(group = hla_class)

## Category 2: Total peptides by class
cat2 <- barplot_data %>%
  group_by(hla_class) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  mutate(cat = 2, pep_type = "all") %>%
  dplyr::rename(group = hla_class)

## Category 3: Split by class and pep_type
cat3 <- barplot_data %>%
  group_by(hla_class, pep_type) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  mutate(cat = 3) %>%
  dplyr::rename(group = hla_class)

## Category 4: Filtered by length and class
cat4 <- barplot_data %>%
  filter(hla_class %in% c("I", "Both"), pep_type != "other") %>%
  group_by(pep_len_cat, pep_type) %>%
  summarise(n = sum(n), .groups = "drop") %>%
  mutate(cat = 4) %>%
  dplyr::rename(group = pep_len_cat)

## Category 5: Known HLA count
cat5 <- hla_df %>%
  distinct(peptide, pep_type) %>%
  dplyr::count(pep_type) %>%
  filter(pep_type != "other") %>%
  mutate(cat = 5, group = "Known_HLA")

## Combine all and calculate unknowns
combined_data <- bind_rows(cat1, cat2, cat3, cat4, cat5) %>%
  pivot_wider(names_from = c(group, cat), values_from = n) %>%
  mutate(Unknown_HLA_5 = `8-12_4` - Known_HLA_5)

## Tidy long format
barplot_data_cat <- combined_data %>%
  pivot_longer(-pep_type, names_to = "name", values_to = "value") %>%
  filter(!is.na(value)) %>%
  mutate(
    cat = as.integer(str_extract(name, "\\d$")),
    group = str_remove(name, "_\\d$")
  ) %>%
  select(-name) %>%
  group_by(cat, pep_type) %>%
  mutate(perc = value / sum(value), sums = sum(value)) %>%
  ungroup()

## Category order
cats <- c("I", "Both", "II", "8-12", "13-35", "Known_HLA", "Unknown_HLA")

## Add color mapping
color_map <- data.frame(
  group = cats,
  fill_color = c(
    blue_col[1], grey_col[5], grey_col[3], blue_col[4], blue_col[2],
    blue_col[7], blue_col[5]
  )
)

## Finalize with colors and factor levels
barplot_data_cat <- barplot_data_cat %>%
  left_join(color_map, by = "group") %>%
  mutate(
    cat = factor(cat, levels = 1:5),
    group = factor(group, levels = cats)
  )

barplot_sums <- barplot_data_cat %>%
  distinct(cat, pep_type, sums) %>%
  mutate(
    rowname = paste0(pep_type, "_", cat),
    sums = formatC(sums, big.mark = ",")
  ) %>%
  column_to_rownames("rowname")

p_overview_1 <- ggplot() +
  geom_bar(
    data = barplot_data_cat %>% filter(cat == 1 | cat == 2),
    aes(x = cat, y = perc, fill = fill_color, group = group),
    stat = "identity"
  ) +
  scale_fill_identity() +
  scale_y_continuous(expand = c(0, 0), name = element_blank()) +
  scale_x_discrete(
    name = element_blank(),
    labels = c(
      paste0(annotations %>%
               distinct(dataset_id) %>%
               nrow(), " datasets"),
      paste0(
        "All peptides\n(n = ",
        barplot_sums["all_2", "sums"], ")"
      )
    ),
    expand = c(0, 0)
  ) +
  theme(
    axis.line = element_blank(), axis.ticks = element_blank(),
    axis.text.y = element_blank()
  )

p_overview_2a <- ggplot() +
  geom_bar(
    data = barplot_data_cat %>% filter(
      (cat == 3 | cat == 4 | cat == 5),
      pep_type == "canonical"
    ),
    aes(x = cat, y = value, fill = fill_color, group = group),
    stat = "identity"
  ) +
  scale_fill_identity() +
  scale_y_continuous(name = element_blank(), expand = c(0, 0)) +
  scale_x_discrete(
    name = element_blank(), expand = c(0, 0),
    labels = c(
      paste0(
        "(n = ",
        barplot_sums["canonical_3", "sums"], ")"
      ),
      paste0(
        "(n = ",
        barplot_sums["canonical_4", "sums"], ")"
      ),
      paste0(
        "(n = ",
        barplot_sums["canonical_5", "sums"], ")"
      )
    )
  ) +
  theme(
    axis.line = element_blank(), axis.ticks = element_blank(),
    axis.text.y = element_blank()
  )

p_overview_2b <- ggplot() +
  geom_bar(
    data = barplot_data_cat %>% filter(
      (cat == 3 | cat == 4 | cat == 5),
      pep_type == "noncanonical"
    ),
    aes(x = cat, y = value, fill = fill_color, group = group),
    stat = "identity"
  ) +
  scale_fill_identity() +
  scale_y_continuous(name = element_blank(), expand = c(0, 0)) +
  scale_x_discrete(
    name = element_blank(), expand = c(0, 0),
    labels = c(
      paste0(
        "(n = ",
        barplot_sums["noncanonical_3", "sums"],
        ")\nHLA class on\nwhich peptide\nwas detected"
      ),
      paste0(
        "(n = ",
        barplot_sums["noncanonical_4", "sums"],
        ")\nPeptide length"
      ),
      paste0(
        "(n = ",
        barplot_sums["noncanonical_5", "sums"],
        ")\n HLA typing\nknown"
      )
    )
  ) +
  theme(
    axis.line = element_blank(), axis.ticks = element_blank(),
    axis.text.y = element_blank()
  )

figure_s6a <- p_overview_1 + p_overview_2a +
  plot_spacer() + p_overview_2b +
  plot_layout(widths = c(2, 2))

ggsave("Figure_S6A.svg",
  device = "svg", path = plot_dir, figure_s6a,
  width = 2000, height = 1300, units = "px"
)
```

## Figure S6B - Classification on protein/ORF level
```{r}
# Link peptides to source ncORF/CDS
pep_pro <- bind_rows(
  pep_can %>%
    semi_join(can_id, "protein_accession") %>%
    distinct(peptide, protein_accession),
  pep_nonc %>% distinct(peptide, protein_accession)
)

# Determine per CDS/ncORF the HLA class(es) on which peptides were detected
protein_hla <- peptide_hla %>%
  distinct(peptide, hla_class) %>%
  inner_join(pep_pro, "peptide", relationship = "many-to-many") %>%
  distinct(protein_accession, hla_class) %>%
  group_by(protein_accession) %>%
  summarise(
    hla_class = ifelse(n_distinct(hla_class) == 2, "Both",
      first(hla_class)
    ),
    .groups = "drop"
  ) %>%
  mutate(is_can = !str_detect(protein_accession, "CONTRIB_GENCODE")) %>%
  dplyr::count(is_can, hla_class) %>%
  mutate(hla_class = factor(hla_class, levels = c("I", "Both", "II")))

figure_s6b1 <- ggplot(
  protein_hla %>% filter(is_can),
  aes(
    x = "Canonical\nproteins", y = n,
    fill = hla_class
  )
) +
  geom_bar(stat = "identity", position = "stack", show.legend = FALSE) +
  scale_x_discrete(expand = c(0, 0), name = "") +
  scale_y_continuous(
    expand = c(0, 0), name = "Number of canonical proteins",
    labels = label_number(scale_cut = cut_short_scale())
  ) +
  scale_fill_manual(values = c(blue_col[1], grey_col[c(5, 3)]))


figure_s6b2 <- ggplot(
  protein_hla %>% filter(!is_can),
  aes(
    x = "Non-canonical\nORFs", y = n,
    fill = hla_class
  )
) +
  geom_bar(stat = "identity", position = "stack") +
  scale_x_discrete(expand = c(0, 0), name = "") +
  scale_y_continuous(
    expand = c(0, 0), name = "Number of non-canonical ORFs",
    labels = comma
  ) +
  scale_fill_manual(
    values = c(blue_col[1], grey_col[c(5, 3)]),
    name = element_blank(),
    labels = c(
      "Detected only in\nHLA-I MS-runs",
      "Detected in both\nHLA-I and HLA-II MS-runs",
      "Detected only in\nHLA-II MS-runs"
    )
  )

figure_s6b <- figure_s6b1 + figure_s6b2
```

## Figure S6C - Classification by peptide length
```{r}
# Count peptides >14 aa and ≤ 14aa per HLA class for ncORF and CDS
pep_class <- peptide_hla %>%
  group_by(peptide, pep_type) %>%
  filter(pep_type != "other") %>%
  summarise(
    hla_class = ifelse(n_distinct(hla_class) == 2, "Both",
      first(hla_class)
    ),
    .groups = "drop"
  ) %>%
  mutate(pep_len = ifelse(nchar(peptide) > 14, ">14 aa", "≤ 14aa")) %>%
  dplyr::count(hla_class, pep_type, pep_len) %>%
  group_by(pep_type) %>%
  mutate(
    perc = n / sum(n) * 100,
    hla_class = factor(hla_class, c("I", "Both", "II"))
  )

figure_s6c <- ggplot(
  pep_class,
  aes(hla_class, perc, group = pep_len, fill = pep_len)
) +
  geom_bar(stat = "identity") +
  scale_x_discrete(
    expand = c(0, 0),
    name = "HLA class on which peptide is detected"
  ) +
  scale_y_continuous(
    expand = c(0, 0), name = "Percentage of all peptides",
    labels = label_percent(scale = 1)
  ) +
  scale_fill_manual(values = blue_col[c(6, 2)], name = "Peptide length") +
  facet_wrap(~pep_type)
```

## Figure S6D - Graph length distribution per dataset
```{r}
# Determine relative length distribution split by HLA class
peptide_lendis <- peptide_hla %>%
  group_by(hla_class, dataset_id) %>%
  mutate(p_count = n(), p_len = nchar(peptide)) %>%
  group_by(dataset_id, hla_class, p_len, p_count) %>%
  summarise(len_count = n(), .groups = "drop") %>%
  mutate(len_count_f = len_count / p_count) %>%
  mutate(dataset_hla = paste0(dataset_id, "__", hla_class))

lendis_n <- peptide_lendis %>%
  distinct(dataset_id, hla_class) %>%
  dplyr::count(hla_class) %>%
  deframe()

figure_s6d <- ggplot(
  data = peptide_lendis %>% filter(p_len <= 30),
  aes(
    x = p_len, y = len_count_f, group = dataset_hla,
    color = hla_class
  )
) +
  geom_line(linewidth = 0.25) +
  geom_point(size = 0.5) +
  scale_y_continuous(
    expand = c(0, 0), limits = c(0, 0.8),
    name = "Relative occurence"
  ) +
  scale_x_continuous(
    expand = c(0, 0), limits = c(6.80, 30),
    name = "Peptide length (aa)"
  ) +
  scale_color_manual(
    values = c(blue_col[1], grey_col[3]), name = "HLA class",
    labels = c(
      paste0(
        "HLA-I (n=", lendis_n["I"],
        " dataset components)"
      ),
      paste0(
        "HLA-II (n=", lendis_n["II"],
        " dataset components)"
      )
    )
  )
```

## Figure S6E - Compare HLA I vs II
```{r}
# Group peptides in those exclusively found in HLA-I or not
orf_hlaclass <- pep_nonc %>%
  mutate(hla_class = ifelse(msrun_hla_classes == "I", "I", "II"))

orf_hlaclass_count <- orf_hlaclass %>%
  dplyr::count(hla_class, orf_biotype) %>%
  group_by(hla_class) %>%
  mutate(maxi = max(n)) %>%
  ungroup() %>%
  mutate(
    # Scale values so that both sides of the axes will be the same length
    scale_factor = max(maxi) / min(maxi),
    orf_biotype = factor(orf_biotype,
      levels = unique(orf_biotype[order(hla_class, n)])
    ),
    n_adj = ifelse(hla_class == "II", n * scale_factor, -n)
  ) %>%
  arrange(hla_class, -n)

class_scale <- orf_hlaclass_count$scale_factor %>% first()

hla_n <- orf_hlaclass_count %>%
  dplyr::count(hla_class, wt = n) %>%
  deframe()

figure_s6e <-
  ggplot(
    orf_hlaclass_count,
    aes(y = orf_biotype, x = n_adj, fill = hla_class)
  ) +
  geom_bar(stat = "identity", position = "identity", show.legend = FALSE) +
  scale_x_continuous(
    position = "top",
    breaks = c(
      -1250, -1000, -750, -500, -250, 0, 20 * class_scale,
      40 * class_scale, 60 * class_scale, 80 * class_scale
    ),
    labels = c(1250, 1000, 750, 500, 250, 0, 20, 40, 60, 80),
    limits = c(-1250, 1250),
    expand = c(0, 0),
    name = "# peptides of ORF biotype detected"
  ) +
  scale_y_discrete(name = element_blank()) +
  scale_fill_manual(
    values = c(blue_col[1], grey_col[5]),
    labels = c(
      paste0("HLA-I (n=", hla_n["I"], " peptides)"),
      paste0("HLA-II (n=", hla_n["II"], " peptides)")
    )
  ) +
  theme(axis.line.y = element_blank(), axis.ticks.y = element_blank())
```

## Figure S6F - Check HLA-II peptides

```{r}
# Determine HLA class(es) per dataset
dataset_class <- annotations %>%
  group_by(dataset_id) %>%
  summarise(hla_class_dataset = ifelse(n_distinct(hla_class) == 2, "Both",
    first(hla_class)
  ))

# Determine per peptide number of detections in HLA-I and HLA-II MS-runs
hlaii_orf_peptides_count <- peptide_hla %>%
  inner_join(
    orf_hlaclass %>% filter(hla_class == "II") %>% distinct(peptide),
    "peptide"
  ) %>%
  dplyr::count(peptide, hla_class, dataset_id) %>%
  inner_join(dataset_class, "dataset_id") %>%
  mutate(len = nchar(peptide)) %>%
  group_by(peptide, hla_class) %>%
  mutate(sum_n = sum(n)) %>%
  group_by(hla_class) %>%
  mutate(maxi = max(sum_n)) %>%
  ungroup() %>%
  mutate(
    # Scale values so that both sides of the axes will be the same length
    scale_factor = max(maxi) / min(maxi),
    n = ifelse(hla_class == "II", n * scale_factor, -n),
    peptide = factor(peptide,
      levels =
        rev(unique(peptide[order(hla_class, -sum_n)]))
    )
  ) %>%
  group_by(peptide, hla_class, hla_class_dataset) %>%
  mutate(color_num = row_number()) %>%
  ungroup() %>%
  mutate(color = case_when(
    hla_class_dataset == "I" ~
      as.character(color_num + 100),
    hla_class_dataset == "II" ~
      as.character(color_num + 200),
    .default = as.character(color_num + 300)
  ))

# Map colors to alternate colors of adjacent bars
hla_color_mapping <- hlaii_orf_peptides_count %>%
  distinct(color) %>%
  arrange(color) %>%
  mutate(
    color_int = as.integer(color),
    color_hex = case_when(
      color_int < 200 & (color_int %% 2 == 0) ~ blue_col[1],
      color_int < 200 & (color_int %% 2 == 1) ~ blue_col[4],
      color_int < 300 & (color_int %% 2 == 0) ~ grey_col[1],
      color_int < 300 & (color_int %% 2 == 1) ~ grey_col[4],
      color_int %% 2 == 0 ~ pink_col[1],
      color_int %% 2 == 1 ~ pink_col[4],
      .default = NA
    )
  )

pep_len_mapping <- hlaii_orf_peptides_count %>%
  arrange(peptide) %>%
  distinct(peptide, len)

hla_ii_scale <- hlaii_orf_peptides_count$scale_factor %>% unique()

detection_count <- hlaii_orf_peptides_count %>%
  distinct(peptide, sum_n, hla_class) %>%
  dplyr::count(hla_class, wt = sum_n) %>%
  mutate(n = formatC(n, big.mark = ",")) %>%
  deframe()

figure_s6f <- ggplot(
  hlaii_orf_peptides_count,
  aes(y = peptide, x = n, fill = color)
) +
  geom_bar(stat = "identity", show.legend = FALSE) +
  scale_fill_manual(
    breaks = hla_color_mapping$color,
    values = hla_color_mapping$color_hex
  ) +
  geom_vline(aes(xintercept = 0)) +
  scale_y_discrete(
    labels = pep_len_mapping$len, guide = guide_axis(n.dodge = 2),
    name = paste0("Peptides (n = ", nrow(pep_len_mapping), ")")
  ) +
  scale_x_continuous(
    position = "top",
    breaks = c(
      -600, -400, -200, 0, 20 * hla_ii_scale,
      40 * hla_ii_scale, 60 * hla_ii_scale
    ),
    labels = c(600, 400, 200, 0, 20, 40, 60),
    limits = c(-60 * hla_ii_scale, 60 * hla_ii_scale),
    expand = c(0, 0),
    name = paste0(
      "Number of peptide detections per dataset\n",
      "On HLA-I (n = ", detection_count["I"], ") ",
      "On HLA-II (n = ", detection_count["II"], ")"
    )
  )
```

## Figure S6B-F - Combine

```{r}
figure_s6bf <- figure_s6b + figure_s6c + figure_s6d + figure_s6e +
  free(figure_s6f) +
  plot_layout(
    design = "ABFFF\nCCFFF\nDDFFF\nEEFFF",
    guides = "collect"
  )

ggsave("figure_S6BF.svg",
  device = "svg", path = plot_dir, figure_s6bf,
  width = 3200, height = 2500, units = "px"
)
```

# Figure S7 - ORF statistics
## Figure S7A-D - ORF overview
```{r}
# Determine percentage detected ncORFs per biotype
orf_pep_stats_ref <- orf_pep_stats %>%
  mutate(is_det = n_pep >= 1) %>%
  dplyr::count(orf_biotype, is_det) %>%
  mutate(mean_perc = sum(n[which(is_det)]) / sum(n)) %>%
  group_by(orf_biotype) %>%
  mutate(
    total = sum(n),
    perc = n / total
  ) %>%
  rowwise() %>%
  mutate(
    conf_low = ifelse(is_det, binom.test(n, total)$conf.int[1], NA),
    conf_high = ifelse(is_det, binom.test(n, total)$conf.int[2], NA),
    p_val = ifelse(is_det, binom.test(n, total, p = mean_perc)$p.value,
      NA
    )
  ) %>%
  ungroup() %>%
  mutate(p_adj = p.adjust(p_val, method = "holm"))

# Determine percentage detected ncORFs per length category
orf_pep_stats_len <- orf_pep_stats %>%
  mutate(is_det = n_pep > 0) %>%
  mutate(len_cat = cut(orf_len,
    breaks = c(15, 30, 50, 75, 100, Inf),
    labels = c("16-30", "31-50", "51-75", "76-100", ">100")
  )) %>%
  dplyr::count(len_cat, is_det) %>%
  group_by(len_cat) %>%
  mutate(total = sum(n)) %>%
  rowwise() %>%
  mutate(
    perc = n / total,
    conf_low = ifelse(is_det, binom.test(n, total)$conf.int[1], NA),
    conf_high = ifelse(is_det, binom.test(n, total)$conf.int[2], NA)
  )

# Determine enrichment in detecteability for longer ncORFs
orf_len_enrichment <- orf_pep_stats_len %>%
  filter(is_det) %>%
  mutate(cat = ifelse(len_cat == "16-30", "low", "high")) %>%
  group_by(cat) %>%
  summarise(perc = sum(n) / sum(total) * 100) %>%
  pivot_wider(names_from = cat, values_from = perc) %>%
  mutate(enrichment = (high - low) / high)

figure_s7a <- ggplot(
  orf_pep_stats_ref,
  aes(fill = is_det, y = n, x = orf_biotype)
) +
  geom_bar(position = "stack", stat = "identity", show.legend = FALSE) +
  scale_x_discrete(name = element_blank(), expand = c(0, 0)) +
  scale_y_continuous(name = "Number of non-canonical ORFs", expand = c(0, 0)) +
  scale_fill_manual(values = green_col[c(4, 1)]) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

figure_s7b <- ggplot(
  orf_pep_stats_ref,
  aes(fill = is_det, y = perc, x = orf_biotype)
) +
  geom_bar(position = "stack", stat = "identity", show.legend = FALSE) +
  scale_x_discrete(name = element_blank(), expand = c(0, 0)) +
  scale_y_continuous(
    name = "Percentage of non-canonical ORFs",
    expand = c(0, 0),
    labels = label_percent()
  ) +
  scale_fill_manual(values = green_col[c(4, 1)]) +
  geom_linerange(aes(ymin = conf_low, ymax = conf_high)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

figure_s7c <- ggplot(
  orf_pep_stats_len,
  aes(fill = is_det, y = n, x = len_cat)
) +
  geom_bar(position = "stack", stat = "identity", show.legend = FALSE) +
  scale_x_discrete(name = "Orf length (aa)", expand = c(0, 0)) +
  scale_y_continuous(name = "Number of non-canonical ORFs", expand = c(0, 0)) +
  scale_fill_manual(values = green_col[c(4, 1)]) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

figure_s7d <- ggplot(
  orf_pep_stats_len,
  aes(fill = is_det, y = perc, x = len_cat)
) +
  geom_bar(position = "stack", stat = "identity", show.legend = FALSE) +
  scale_x_discrete(name = "Orf length (aa)", expand = c(0, 0)) +
  scale_y_continuous(
    name = "Percentage of non-canonical ORFs", expand = c(0, 0),
    labels = label_percent()
  ) +
  scale_fill_manual(values = green_col[c(4, 1)]) +
  geom_linerange(aes(ymin = conf_low, ymax = conf_high)) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

figure_s7ad <- figure_s7a + figure_s7b + figure_s7c + figure_s7d +
  plot_layout(nrow = 1)

ggsave("Figure_S7AD.svg",
  device = "svg", path = plot_dir, figure_s7ad,
  width = 2340, height = 1000, units = "px"
)
```

## Figure S7E - Compare cancer vs non-cancer with cancer genes
```{r}
cancer_census <- read_csv(cancer_census_file)
colnames(cancer_census) <- str_replace_all(colnames(cancer_census), " ", "_")

# Merge orf data with cancer census data
orf_gene <- orf_sequences %>%
  select(-protein) %>%
  left_join(
    cancer_census %>%
      distinct(Gene_Symbol, Role_in_Cancer) %>%
      mutate(is_cancergene = TRUE),
    c("gene_name" = "Gene_Symbol")
  )

# Determine per peptide whether it is found in cancer samples or originates
#  from a cancer gene
orf_cancer <- peptide_hla %>%
  inner_join(pep_nonc, "peptide", relationship = "many-to-many") %>%
  left_join(orf_gene, c("orf_name", "orf_biotype")) %>%
  distinct(peptide, is_cancer, orf_biotype, is_cancergene) %>%
  group_by(peptide, orf_biotype, is_cancergene) %>%
  summarise(
    is_cancer = ifelse(n() > 1, "Both", first(is_cancer)),
    .groups = "drop"
  ) %>%
  mutate(
    is_cancergene = ifelse(is.na(is_cancergene), FALSE, TRUE),
    is_cancer = factor(is_cancer,
      levels = c("Cancer", "Both", "Non-cancer")
    )
  )

# Determine counts per cancercategory and orf biotype
orf_cancer_count <- orf_cancer %>%
  dplyr::count(is_cancer, orf_biotype, is_cancergene) %>%
  group_by(is_cancer) %>%
  mutate(
    orf_biotype = str_replace_all(orf_biotype, " ", "\n"),
    total = sum(n),
    perc = n / total * 100,
    orf_biotype = factor(orf_biotype,
      levels = unique(orf_biotype[order(is_cancer, -perc)])
    )
  ) %>%
  group_by(is_cancer, orf_biotype) %>%
  mutate(sumn = sum(n)) %>%
  rowwise() %>%
  mutate(
    conf_low = ifelse(!is_cancergene,
      binom.test(sumn, total)$conf.int[1] * 100, NA
    ),
    conf_high = ifelse(!is_cancergene,
      binom.test(sumn, total)$conf.int[2] * 100, NA
    ),
    color = case_when(
      is_cancergene ~ red_col,
      is_cancer == "Cancer" ~ blue_col[2],
      is_cancer == "Non-cancer" ~ blue_col[6],
      TRUE ~ grey_col[4]
    ) %>% factor(levels = c(blue_col[2], blue_col[6], grey_col[4], red_col)),
    x_axis = paste0(orf_biotype, is_cancer)
  ) %>%
  arrange(orf_biotype, is_cancer) %>%
  mutate(x_axis = factor(x_axis, levels = unique(x_axis)))

cancer_x_labels <- orf_cancer_count %>%
  filter(is_cancer == "Both") %>%
  arrange(orf_biotype) %>%
  distinct(x_axis, orf_biotype)

cancer_counts <- orf_cancer_count %>%
  ungroup() %>%
  distinct(is_cancer, total) %>%
  mutate(total = paste0("(n = ", total, " peptides)")) %>%
  deframe()

cancergene_count <- orf_cancer_count %>%
  filter(is_cancergene) %>%
  pull(n) %>%
  sum() %>%
  paste0("(n = ", ., " peptides)")

figure_s7e <-
  ggplot(orf_cancer_count, aes(x = x_axis, y = perc, fill = color)) +
  geom_bar(stat = "identity", position = "stack") +
  scale_fill_identity(
    guide = "legend",
    breaks = c(blue_col[2], grey_col[4], blue_col[6], red_col),
    labels = c(
      paste0("Cancer\n", cancer_counts["Cancer"]),
      paste0("Both\n", cancer_counts["Both"]),
      paste0("Non-cancer\n(", cancer_counts["Non-cancer"]),
      paste0("Peptide from ORF\nlocated at cancergene\n", cancergene_count)
    ),
    name = "Sample type in which\npeptide was detected"
  ) +
  scale_y_continuous(
    expand = c(0, 0), name = "% of ORF biotype detected",
    labels = label_percent(scale = 1)
  ) +
  scale_x_discrete(
    expand = c(0, 0), name = element_blank(),
    breaks = cancer_x_labels$x_axis,
    labels = cancer_x_labels$orf_biotype
  ) +
  facet_grid(~orf_biotype, scales = "free") +
  geom_linerange(aes(ymin = conf_low, ymax = conf_high)) +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    strip.text.x = element_blank()
  )

ggsave("Figure_S7E.svg",
  device = "svg", path = plot_dir, figure_s7e,
  width = 1400, height = 1200, units = "px"
)
```

## Figure S7F - Compare on peptide level
```{r}
neworf_cancer <- orf_cancer %>%
  distinct(peptide, is_cancer) %>%
  # Rename for function get_stats
  dplyr::rename(protein = peptide) %>%
  get_stats() %>%
  dplyr::rename(peptide = protein) %>%
  mutate(weight = mw(peptide) / len)

cancer_colors <- c(blue_col[2], grey_col[4], blue_col[6])

figure_s7f1 <- create_dis_plot(
  ggplot(
    neworf_cancer,
    aes(x = len, color = is_cancer)
  ),
  "Peptide length", cancer_colors
)

figure_s7f2 <- create_dis_plot(
  ggplot(
    neworf_cancer,
    aes(x = weight, color = is_cancer)
  ),
  "Mean mass per amino acid (u)", cancer_colors,
  y_name = element_blank()
)

figure_s7f3 <- create_dis_plot(
  ggplot(
    neworf_cancer,
    aes(x = hydro, color = is_cancer)
  ),
  "Hydrophobicity (Kyte-Doolittle)", cancer_colors
)

figure_s7f4 <- create_dis_plot(
  ggplot(
    neworf_cancer,
    aes(x = pi, color = is_cancer)
  ),
  "Isoelectric point", cancer_colors,
  y_name = element_blank()
)

figure_s7f <- wrap_plots(
  list(figure_s7f1, figure_s7f2, figure_s7f3, figure_s7f4),
  ncol = 2, nrow = 2
)

ggsave("Figure_S7F.svg",
  device = "svg", path = plot_dir, figure_s7f,
  width = 1200, height = 1200, units = "px"
)
```

# Figure S8 - Hydrophobicity and expression
## Figure S8C - Hydrophobicity at ncORF ends
```{r}
h_lab <- "Hydrophobicity\n(Kyte-Doolittle)"
h_cats <- c("Non-canonical__start", "Non-canonical__end")

# Get p-values for comparisons adjusten using Holm-Bonferoni method
p_vals_start_end <-
  inner_join(
    seq_stats_properties %>%
      filter(pro_cat == h_cats[1]) %>%
      select(hydro, detected, protein_accession) %>%
      dplyr::rename(start = hydro),
    seq_stats_properties %>%
      filter(pro_cat == h_cats[2]) %>%
      select(hydro, detected, protein_accession) %>%
      dplyr::rename(end = hydro),
    c("protein_accession", "detected")
  ) %>%
  select(-protein_accession) %>%
  get_pvals_holms() %>%
  arrange(desc(name))

figure_s8c_plots <- lapply(1:2, function(i) {
  cat <- h_cats[i]
  plot_data <- seq_stats_properties %>%
    filter(pro_cat == cat) %>%
    mutate(y_val = hydro)
  p_val <- p_vals_start_end$value
  create_violin(plot_data, h_lab, p_val, c(-4.1, 2.5),
    n_colors = c((4 * i - 3):(4 * i - 2))
  )
})

figure_s8c_plots[[2]] <- figure_s8c_plots[[2]] + remove_y
figure_s8c <- wrap_plots(figure_s8c_plots, ncol = 2)

ggsave("Figure_S8C.svg",
  device = "svg", path = plot_dir, figure_s8c,
  width = 600, height = 300, units = "px"
)
```

## Figure S8D - Hydrophobicity per biotype
```{r}
# Group seq_stats_hydro by CDS and ncORF biotype
seq_stats_hydro_biotype <- seq_stats_hydro %>%
  group_by(orf_biotype, c_offset, detected) %>%
  mutate(
    low_conf = t.test(hydro)$conf.int[1],
    high_conf = t.test(hydro)$conf.int[2]
  ) %>%
  summarise(
    hydro = mean(hydro), low_conf = first(low_conf),
    high_conf = first(high_conf)
  ) %>%
  mutate(orf_biotype = factor(orf_biotype,
    levels = c(
      "CDS", "uORFs", "uoORFs", "intORFs",
      "doORFs", "dORFs", "lncRNA ORFs",
      "Processed transcript ORFs"
    )
  ))

figure_s8d <- ggplot(
  seq_stats_hydro_biotype,
  aes(x = c_offset, y = hydro, group = detected)
) +
  geom_linerange(aes(ymin = low_conf, ymax = high_conf, color = detected),
    alpha = 0.5, linewidth = 0.3
  ) +
  geom_point(mapping = aes(color = detected), size = .2) +
  geom_smooth(
    method = "loess", formula = y ~ x, se = FALSE,
    mapping = aes(color = detected), linewidth = .7
  ) +
  scale_x_continuous(
    expand = c(0, 0),
    name = "Position relative to C-terminus"
  ) +
  scale_y_continuous(
    limits = c(-3.8, 2.6), expand = c(0, 0),
    name = "Hydrophobicity\n(Kyte-Doolittle)"
  ) +
  scale_color_manual(values = c(green_col[1], blue_col[2])) +
  facet_wrap(~orf_biotype, ncol = 4)

ggsave("figure_S8D.svg",
  device = "svg", path = plot_dir, figure_s8d,
  width = 2516.16, height = 900, units = "px"
)
```

## Figure S8E - Expression per biotype
```{r}
orf_gtex_split <- orf_gtex %>%
  make_biotype_levels() %>%
  split(~orf_biotype)

gtex_violins_pvals <- sapply(names(orf_gtex_split), function(i) {
  p_val <- perform_wtest(orf_gtex_split[[i]])
}) %>%
  p.adjust(method = "holm") %>%
  sapply(format_pval)

gtex_violins <- lapply(names(orf_gtex_split), function(i) {
  p_val <- gtex_violins_pvals[i]
  expression_violin <- create_violin(orf_gtex_split[[i]],
    "Mean FPKM of gene\nexpressing ncORF in GTEx", p_val, c(0, 1600),
    n_colors = c(3, 4), pseudo_log = TRUE, x_label = i,
    y_breaks = c(0, 10, 250, 500, 1000, 2000)
  )
})

gtex_violins_f <- lapply(1:7, function(i) {
  one_plot <- gtex_violins[[i]]
  if (i > 1) {
    one_plot <- one_plot + remove_y
  }
  return(one_plot)
})

figure_s8e <- wrap_plots(gtex_violins_f, nrow = 1)

ggsave("figure_S8E.svg",
  device = "svg", path = plot_dir, figure_s8e,
  width = 2300, height = 600, units = "px"
)
```

## Figure S8F - Expression per tissue
```{r}
# Process GTEX data to longer format with improved names
gtex_tissue <- gtex_data %>%
  dplyr::rename(gene_id = X) %>%
  pivot_longer(-gene_id) %>%
  dplyr::rename(tissue_raw = name, fpkm = value) %>%
  mutate(
    tissue = str_replace_all(str_extract(tissue_raw, "^\\w+"), "_", " "),
    tissue = case_when(
      tissue == "Adipose" ~ "Adipose Tissue",
      tissue == "Artery" ~ "Blood Vessel",
      tissue_raw == "Cells.EBV_transf.lymphocytes" ~ "Blood",
      tissue_raw == "Cells.transformed_fibroblasts" ~ "Skin",
      tissue == "FallopianTube" ~ "Fallopian Tube",
      tissue == "Minor Salivary Gland" ~ "Salivary Gland",
      tissue == "Whole Blood" ~ "Blood",
      TRUE ~ tissue
    )
  ) %>%
  group_by(tissue, gene_id) %>%
  summarise(fpkm = mean(fpkm))

# Combine GTEX expression per biotype with ncORF genes
orf_gtex_tissue <-
  seq_stats %>%
  inner_join(orf_sequences %>%
               select(protein_accession, gene_id), "protein_accession") %>%
  inner_join(gtex_tissue, "gene_id", relationship = "many-to-many")

# Statistical test per tissue, print highest p-value for reporting
p_tissue <- orf_gtex_tissue %>%
  mutate(y_val = fpkm) %>%
  split(~tissue) %>%
  sapply(perform_wtest) %>%
  p.adjust(method = "holm")
max(p_tissue)

figure_s8f <- ggplot(orf_gtex_tissue, aes(detected, fpkm, color = detected)) +
  geom_boxplot(outliers = FALSE) +
  scale_y_continuous(
    trans = scales::pseudo_log_trans(base = 10),
    breaks = c(0, 2, 10, 50, 250),
    name = "Mean FPKM of gene\nexpressing ncORF in GTEx"
  ) +
  scale_x_discrete(labels = c(), name = element_blank()) +
  facet_wrap(~tissue, nrow = 1) +
  scale_color_manual(values = green_col[c(1, 4)])

ggsave("Figure_S8F.svg",
  device = "svg", path = plot_dir, figure_s8f,
  width = 2300, height = 600, units = "px"
)
```
